---
title: "100-time-tree"
output: html_document
date: "2023-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
library(tidyverse)
library(ggtree)
library(phytools)
library(ggrepel)
library(ggpubr)
library(viridis)
```

Generate taxon list

```{r}
meta<-read.csv("data/loci.csv")
meta
```

```{r}
organisms<-meta %>% select(organism) %>% unique()
organisms$organism<-gsub("_"," ",organisms$organism)
organisms
write_tsv(organisms, col_names = FALSE, file="outputs/100/species-list-unedited.tsv")
```

79 Mismatches

Thylacine cynocephalus (not found in NCBI taxonomy)
Equus ferus thistlecreek (not found in NCBI taxonomy)
Nannopterum brasilianus (not found in NCBI taxonomy)
Scyliorhinus torazame (replaced with Carcharhinus leucas)
Guentherus altivela (replaced with Ateleopus japonicus)
Paramormyrops kingsleyae (replaced with Paramormyrops curvifrons)
Phrynoglossus myanhessei (replaced with Occidozyga laevis)
Bombina variegata variegata (replaced with Bombina variegata)
Bettongia penicillata ogilbyi (replaced with Bettongia penicillata)
Solenodon paradoxus woodi (replaced with Solenodon paradoxus)
Prionailurus bengalensis euptilurus (replaced with Prionailurus bengalensis)
Mustela putorius furo (replaced with Mustela putorius)
Ailurus fulgens (replaced with Ailurus)
Ursus arctos horribilis (replaced with Ursus arctos)
Equus asinus somalicus (replaced with Equus asinus)
Equus africanus asinus (replaced with Equus asinus)
Equus burchellii boehmi (replaced with Equus quagga)
Equus zebra hartmannae (replaced with Equus zebra)
Dicerorhinus sumatrensis sumatrensis (replaced with Dicerorhinus sumatrensis)
Ceratotherium simum cottoni (replaced with Ceratotherium simum)
Lama guanicoe cacsilensis (replaced with Lama guanicoe)
Neophocaena asiaeorientalis (replaced with Neophocaena)
Kobus leche leche (replaced with Kobus leche)
Hippotragus niger niger (replaced with Hippotragus niger)
Cervus hanglu yarkandensis (replaced with Cervus hanglu)
Cervus elaphus hippelaphus (replaced with Cervus elaphus)
Pteronotus parnellii (replaced with Pteronotus davyi)
Murina aurata feae (replaced with Murina aurata)
Peromyscus maniculatus bairdii (replaced with Peromyscus maniculatus)
Peromyscus polionotus subgriseus (replaced with Peromyscus polionotus)
Colobus angolensis palliatus (replaced with Colobus angolensis)
Tupaia chinensis (replaced with Tupaia glis)
Notechis scutatus (replaced with Notechis)
Pituophis catenifer pumilus (replaced with Pituophis catenifer)
Diadophis punctatus similis (replaced with Diadophis punctatus)
Vipera berus berus (replaced with Vipera berus)
Varanus salvator macromaculatus (replaced with Varanus salvator)
Elgaria multicarinata webbii (replaced with Elgaria multicarinata)
Nothoprocta perdicaria (replaced with Nothoprocta)
Tinamus guttatus (replaced with Tinamus)
Apteryx australis mantelli (replaced with Apteryx mantelli)
Chaetura pelagica (replaced with Chaetura)
Antrostomus carolinensis (replaced with Lyncornis macrotis)
Egretta garzetta (replaced with Egretta novaehollandiae)
Pelecanus crispus (replaced with Pelecanus occidentalis)
Limosa lapponica baueri (replaced with Limosa lapponica)
Recurvirostra avosetta (replaced with Recurvirostra)
Phoenicopterus ruber (replaced with Phoenicopterus)
Merops nubicus (replaced with Merops apiaster)
Colius striatus (replaced with Colius)
Strix occidentalis caurina (replaced with Strix uralensis)
Athene cunicularia (replaced with Athene)
Aquila chrysaetos canadensis (replaced with Aquila chrysaetos)
Phylloscopus trochiloides viridanus (replaced with Phylloscopus trochiloides)
Corvus cornix cornix (replaced with Corvus cornix)
Lonchura striata domestica (replaced with Lonchura striata)
Mesoclemmys tuberculata (replaced with Phrynops hilarii)
Malaclemys terrapin terrapin (replaced with Malaclemys terrapin)
Trachemys scripta (replaced with Trachemys dorbigni)
Cebus capucinus imitator (no substitute found)
Eulemur fulvus (no substitute found)
Eidolon dupreanum (no substitute found)
Gopherus evgoodei (no substitute found)
Cercopithecus albogularis (no substitute found)
Canis lupus dingo (no substitute found)
Pteropus pselaphon (no substitute found)
Eurynorhynchus pygmeus (no substitute found)
Oryzias melastigma (no substitute found)
Psittacula krameri (no substitute found)
Eptesicus fuscus (no substitute found)
Hipposideros pendleburyi (no substitute found)
Phalacrocorax harrisi (no substitute found)
Phalacrocorax carbo (no substitute found)
Urile pelagicus (no substitute found)
Miniopterus schreibersii (no substitute found)
Ardeotis arabs (no substitute found)
Bubo blakistoni (no substitute found)
Anabarilius grahami (no substitute found)
Nannopterum harrisi (no substitute found)


## Moving forward without resolving names

```{r}
tree
```
```{r}
data<-meta 
data$Species<-gsub("_"," ",data$organism)
data<-as_tibble(data)
data$locus_name<-data$assigned_name
list<-split(data, f=data$locus_name)
```


```{r}
writeList<-function(table) {
  species<-as_tibble(unique(table$Species))
  if(length(species$value) > 1) {
  write_tsv(species, col_names = FALSE, paste0("outputs/100/lists/",unique(table$locus_name),".tsv"))
  }
}
```

This takes a while, so commenting out.     

```{r, eval=FALSE}
lapply(list, writeList)
```


## Tree   

```{r}
tree<-read.tree("outputs/100/species-list-unedited.nwk")
t<-ggtree(tree)
```

```{r}
#Making tip ages = 0 
t$data<-t$data %>% mutate(Age=max(x)-x)

t + geom_tiplab(size=2) +
  geom_treescale()+
  xlim(0,600)+
  geom_nodelab(aes(label=round(Age,2)), hjust=0, size=2)
```

```{r}
#Making "Taxa" for tip labs and calculate number of taxa
#Exlude unassigned "NK" locus_names
#filter to tips in tree
treetips<-t$data %>% filter(isTip==TRUE) %>% select(label)

df<-data %>% filter(locus_name != "NK") %>% unique() %>%
  mutate(Delim=gsub(" ","_",Species)) %>%
  filter(Delim %in% treetips$label) %>%
  group_by(locus_name) %>% mutate(Taxa=paste(Delim, collapse=" "))  %>% mutate(Count=n())

#getting min age of nodes
df2<-df %>% mutate(node = as.numeric(ifelse(Count > 1, findMRCA(tree, tips=strsplit(Taxa, " ")[[1]]), NA)))
df3<-left_join(df2, t$data, by="node")
df3$Age<-replace_na(df3$Age, 0)


#Summarize min age of nodes
df3 %>% select(locus_name, Age) %>% unique()
```

```{r}
annotations<-df3 %>% select(locus_name, node, Age, x, y, host_class, virus_family, virus_genus) %>% unique()

minAges<-annotations %>% rename(MinimumAge=Age) %>% select(virus_genus, locus_name, MinimumAge, host_class)

write_csv(minAges, "outputs/100/min-ages.csv")
```


Annotating tree

```{r}
fact<-8
bigdf<-left_join(data, t$data, by=c("organism"="label"))  %>%
  drop_na() %>% #Some species not in timetree, subspecies added to newick string where appropriate
  select(Species, organism, virus_family, virus_genus, x, y, locus_name) %>%
  unique() %>% select(-locus_name) %>% 
  group_by(Species, organism, virus_family, virus_genus, x, y) %>%
  summarize(Count=n()) %>%
  unique() %>%
  mutate(newX = ifelse(virus_family=="Bornaviridae", x+fact,
                ifelse(virus_family=="Hepadnaviridae", x+2*fact,
                ifelse(virus_family=="Parvoviridae", x+3*fact,
                ifelse(virus_family=="Circoviridae", x+4*fact,
                ifelse(virus_family=="Filoviridae", x+5*fact,
                ifelse(virus_family=="Chuviridae", x+6*fact,
                ifelse(virus_family=="Paramyxoviridae", x+7*fact,
                ifelse(virus_family=="Flaviviridae", x+8*fact,
                ifelse(virus_family=="Retroelement", x+9*fact))))))))))

bigdf$virus_family<-factor(bigdf$virus_family, levels=c( "Hepadnaviridae","Circoviridae", "Parvoviridae","Retroelement","Flaviviridae",
                                                         "Bornaviridae","Chuviridae","Filoviridae","Paramyxoviridae"))
```

```{r}
nodes<-filter(annotations, node != "NA")

nodes$virus_family<-factor(nodes$virus_family, levels=c( "Hepadnaviridae","Circoviridae", "Parvoviridae","Retroelement","Flaviviridae",
                                                         "Bornaviridae","Chuviridae","Filoviridae","Paramyxoviridae"))
tips<-filter(annotations, Age == 0)
tipSpecies<-df3 %>% filter(locus_name %in% tips$locus_name)


v<-t + xlim(0,600)+
  geom_point(data = nodes, aes(x=x, y=y, fill=virus_family), size=4, alpha=0.75, pch=23, position = "jitter") +
  geom_point(data= bigdf, aes(x=newX, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" ) 

v

ggsave("outputs/100/starting-tree.pdf", width=12, height=24)
```