---
title: "102-actual-final-tree"
output: html_document
date: "2023-09-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggtree)
library(phytools)
library(ggrepel)
library(ggpubr)
library(viridis)
```

Generate taxon list

Correcting names from outputs/100/unresolved-names.txt
Dropping those without NCBI taxonomy, but otherwise matching.    

```{r}
meta<-read_csv("data/actual-final-vertebrate-eves.csv")
meta
```

```{r}
organisms<-meta %>% dplyr::select(organism) %>% unique()
organisms$organism<-gsub("_"," ",organisms$organism)
write_tsv(organisms, col_names = FALSE, file="outputs/102/species-list-unedited.tsv")

#toedit<-organisms %>% mutate(EditedName=organism)
#write_tsv(toedit, col_names = FALSE, file="outputs/100/species-list-edited.tsv")
edited<-read_tsv("outputs/102/species-list-edited.tsv", col_names = c("Species","edited"))
write_tsv(edited %>% select(edited), col_names = FALSE, file="outputs/102/species-list-edited-for-timetree.tsv")
```

24 unresolved names     
Equus ferus thistlecreek (not found in NCBI taxonomy)
Nannopterum brasilianus (not found in NCBI taxonomy)
Thylacine cynocephalus (not found in NCBI taxonomy)
Equus burchellii boehmi (replaced with Equus quagga)
Cebus capucinus imitator (no substitute found)
Eidolon dupreanum (no substitute found)
Ardeotis arabs (no substitute found)
Phalacrocorax carbo (no substitute found)
Eulemur fulvus (no substitute found)
Bubo blakistoni (no substitute found)
Eptesicus fuscus (no substitute found)
Phalacrocorax harrisi (no substitute found)
Pteropus pselaphon (no substitute found)
Urile pelagicus (no substitute found)
Eurynorhynchus pygmeus (no substitute found)
Psittacula krameri (no substitute found)
Cercopithecus albogularis (no substitute found)
Anabarilius grahami (no substitute found)
Oryzias melastigma (no substitute found)
Hipposideros pendleburyi (no substitute found)
Canis lupus dingo (no substitute found)
Miniopterus schreibersii (no substitute found)
Gopherus evgoodei (no substitute found)
Nannopterum harrisi (no substitute found)

(note, when redoing this on 08212023 I still had 709 unique names)
(note, when redoing this on 09202023 I added 
Acipenser ruthenus	Acipenser ruthenus      
Sciurus vulgaris	Sciurus vulgaris       
Dendropsophus ebraccatus	Dendropsophus ebraccatus)

## Moving forward without resolving names

```{r}
data<-meta 
data$Species<-gsub("_"," ",data$organism)
data<-as_tibble(data) %>% left_join(edited)
data$Species<-data$edited
data$locus_name<-data$assigned_name
list<-split(data, f=data$locus_name)
data$Species<-gsub(" ","_",data$Species)
```

Species column has edited names. 
```{r}
writeList<-function(table) {
  species<-as_tibble(unique(table$Species))
  if(length(species$value) > 1) {
  write_tsv(species, col_names = FALSE, paste0("outputs/102/lists/",unique(table$locus_name),".tsv"))
  }
}
```

This takes a while, so not evaluating unless intending to.     

```{r, eval=FALSE}
lapply(list, writeList)
```


## Tree   

```{r}
#tree<-read.tree("outputs/100/species-list-unedited.nwk")
tree<-read.tree("outputs/102/species-list-edited-for-timetree.nwk")
t<-ggtree(tree, color="grey50")
```

```{r}
#Making tip ages = 0 
t$data<-t$data %>% mutate(Age=max(x)-x)

t + geom_tiplab(size=2) +
  geom_treescale()+
  xlim(0,600)+
  geom_nodelab(aes(label=round(Age,2)), hjust=0, size=2)
```

```{r}
#Making "Taxa" for tip labs and calculate number of taxa
#Exlude unassigned "NK" locus_names
#filter to tips in tree
treetips<-t$data %>% filter(isTip==TRUE) %>% select(label)

df<-data %>% filter(locus_name != "NK") %>% unique() %>%
  mutate(Delim=gsub(" ","_",Species)) %>%
  filter(Delim %in% treetips$label) %>%
  group_by(locus_name) %>% mutate(Taxa=paste(Delim, collapse=" "))  %>% mutate(Count=n()) %>% #Ned to find loci occurring in > 1 species, not just > 1 in a species 
  group_by(locus_name) %>% mutate(TaxonCount=length(unique(Species))) %>% ungroup() %>%
  group_by(locus_name)



#getting min age of nodes
df2<-df %>% mutate(node = as.numeric(ifelse(Count > 1, findMRCA(tree, tips=strsplit(Taxa, " ")[[1]]), NA)))
df3<-left_join(df2, t$data, by="node")
df3$Age<-replace_na(df3$Age, 0)

df3 <- df3 %>% mutate(NewAge=ifelse(TaxonCount > 1, Age, 0))

#Summarize min age of nodes
df3 %>% select(locus_name, Age, NewAge) %>% unique() %>% arrange(-Age, NewAge)

df3<-df3 %>% select(-Age) %>% rename(Age=NewAge)
#Check to make sure some loci with >1 occurrence in 1 organism don't have duplicates treated as loci with old ages
df %>% filter(Species=="Eptatretus_burgeri") %>% select(Species, Count, TaxonCount)
df3 %>% filter(Species=="Eptatretus_burgeri") %>% select(Species, Count, TaxonCount, Age)
```

```{r}
annotations<-df3 %>% select(locus_name, node, Age, x, y, tax_class, virus_family, virus_genus) %>% unique()

minAges<-annotations %>% rename(MinimumAge=Age) %>% select(virus_genus, locus_name, MinimumAge, tax_class)

write_csv(minAges, "outputs/102/min-ages.csv")
write_csv(minAges %>% arrange(-MinimumAge), "tree-figure/min-ages.csv")

minAges %>% arrange(-MinimumAge)
```


Annotating tree

```{r}
fact<-8
bigdf<-left_join(data, t$data, by=c("Species"="label"))  %>%
  drop_na() %>% #Some species not in timetree, subspecies added to newick string where appropriate
  select(Species, organism, virus_family, virus_genus, x, y, locus_name) %>%
  unique() %>% select(-locus_name) %>% 
  group_by(Species, organism, virus_family, virus_genus, x, y) %>%
  summarize(Count=n()) %>%
  unique() %>%
   mutate(newX = ifelse(virus_family=="Alloherpesviridae", x+fact,
                ifelse(virus_family=="Herpesviridae", x+2*fact,
                ifelse(virus_family=="Bornaviridae", x+3*fact,
                ifelse(virus_family=="Chuviridae", x+4*fact,
                ifelse(virus_family=="Circoviridae", x+5*fact,
                ifelse(virus_family=="Filoviridae", x+6*fact,
                ifelse(virus_family=="Flaviviridae", x+7*fact,
                ifelse(virus_family=="Hepadnaviridae", x+8*fact,
                ifelse(virus_family=="Paramyxoviridae", x+9*fact,
                ifelse(virus_family=="Parvoviridae", x+10*fact))))))))))) %>%
  filter(virus_family != "Retroelement")

bigdf$virus_family<-factor(bigdf$virus_family, levels=c( "Alloherpesviridae","Herpesviridae", "Bornaviridae","Chuviridae","Circoviridae",
                                                         "Filoviridae","Flaviviridae","Hepadnaviridae","Paramyxoviridae","Parvoviridae"))

bigdf
```

```{r}
nodes<-filter(annotations %>% select(-tax_class) %>% unique(), node != "NA")

nodes$virus_family<-factor(nodes$virus_family, levels=c( "Alloherpesviridae","Herpesviridae", "Bornaviridae","Chuviridae","Circoviridae",
                                                         "Filoviridae","Flaviviridae","Hepadnaviridae","Paramyxoviridae","Parvoviridae"))

tips<-filter(annotations, Age == 0)
tipSpecies<-df3 %>% filter(locus_name %in% tips$locus_name)


v<-t + xlim(0,600)+
  geom_point(data = nodes, aes(x=x, y=y, fill=virus_family), size=3, alpha=0.75, pch=23, position = "jitter") +
  geom_point(data= bigdf, aes(x=newX, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" ) 

v

ggsave("tree-figure/starting-tree.pdf", width=12, height=24)
```


## Label clades

```{r}
ddf<-data
dddf<-ddf %>% filter(organism %in% tree$tip.label)

findMeANode<-function(Taxon) {
  species <- dplyr::filter(dddf, tax_class == Taxon)
  speciesNode <- findMRCA(tree, tips=species$organism)
  return(speciesNode)
}

findMeANode("Aves")
```

```{r}
classes<-dddf %>% ungroup %>% select(tax_class) %>% unique()
cvector<-dddf$tax_class %>% unique()

cnodes<-lapply(cvector, findMeANode)
```

```{r}
cnodes[sapply(cnodes, is.null)] <- NA

n<-as_tibble(as.vector(unlist(cnodes)))

cladelabels<-bind_cols(classes,n) %>% filter(value != "NA") %>% unique() %>% filter(tax_class!="Unclassified")
cladetips<-bind_cols(classes,n)
cladetips<-cladetips[is.na(cladetips$value),]

ctips<-dddf  %>% filter(tax_class %in% cladetips$tax_class) %>% rename(label=Species) %>% left_join(t$data)

nn<-cladelabels$value
ll<-cladelabels$tax_class
off<-42
```


```{r}
x<-t + xlim(0,600)+
  geom_point(data = nodes %>% filter(Age !=0), aes(x=x, y=y, fill=virus_family), size=3, alpha=0.75, pch=23, position = "jitter") +
  geom_point(data= bigdf, aes(x=newX, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" )+
  geom_text(data=ctips, aes(y=y, x=x+9*fact, label=tax_class, hjust=0)) +
    geom_cladelabel(nn[1],ll[1], offset=off) +
    geom_cladelabel(nn[2],ll[2], offset=off) +
    geom_cladelabel(nn[3],ll[3], offset=off) +
    geom_cladelabel(nn[4],ll[4], offset=off) +
    geom_cladelabel(nn[5],ll[5], offset=off) +
    geom_cladelabel(nn[6],ll[6], offset=off) 

x

ggsave("outputs/102/starting-tree-with-clade-labels.pdf", width=12, height=24)

```


node label version    

```{r}
nnn<-cbind(nn,ll) %>% as_tibble()
nnn$nn<-as.integer(nnn$nn)
cnodes<-t$data %>% filter(node %in% cnodes) %>% left_join(nnn, by=c("node"="nn"))
cnodes
```

```{r}
singles<-ctips %>% select(parent, tax_class) %>% rename(single=parent) %>% left_join(t$data, by=c("single"="node"))
singles
```


```{r}
y<-t + xlim(0,600)+
  geom_point(data = nodes %>% filter(Age !=0), aes(x=x, y=y, fill=virus_family), size=3, alpha=0.75, pch=23, position = "jitter") +
  geom_point(data= bigdf, aes(x=newX, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" ) +
  geom_label_repel(data=cnodes, aes(x=x, y=y, label=ll), max.overlaps = Inf) +
  geom_label_repel(data=singles, aes(x=x, y=y, label=tax_class), max.overlaps = Inf)

y
ggsave("tree-figure/starting-tree-with-node-labels.pdf", width=12, height=24)

```


Make nodes proportional to occurrences
```{r}
summed<-nodes %>% filter(Age !=0) %>% ungroup() %>% group_by(virus_family, node, Age, x, y) %>% summarize(Count=n())

u<-ggtree(tree, color="grey50", size=0.25)

```

```{r}
z<-u + xlim(0,600)+
  geom_point(data = summed, aes(x=x, y=y, fill=virus_family, size=Count), alpha=0.75, pch=23, position="jitter") +
  geom_point(data= bigdf, aes(x=newX, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" ) +
#  geom_label_repel(data=cnodes, aes(x=x, y=y, label=ll), max.overlaps = Inf) +
#  geom_label_repel(data=singles, aes(x=x, y=y, label=tax_class), max.overlaps = Inf) +
  ylim(1,max(t$data$y)+1)

z
ggsave("tree-figure/starting-tree-summed-nodes.pdf", width=8.5, height=11)

```

```{r}
u + xlim(0,0)+
  geom_point(data = summed, aes(x=x, y=y, fill=virus_family, size=Count), alpha=0.75, pch=23, position="jitter") +
  geom_point(data= bigdf, aes(x=x+50, y=y, fill=virus_family, size=Count), alpha=0.75, pch=21) +
  theme_tree2()+
  theme(axis.title.x = element_text(size=14)) +
  theme(legend.position = "right") +
  scale_x_continuous(name="", breaks=c(063, 263, 463),
                     labels=c(500,300,100) )+
  #coord_cartesian(xlim = c(373,550)) + #was 323
  scale_fill_viridis_d(option = "H", name="Virus Family" ) +
#  geom_label_repel(data=cnodes, aes(x=x, y=y, label=ll), max.overlaps = Inf) +
#  geom_label_repel(data=singles, aes(x=x, y=y, label=tax_class), max.overlaps = Inf) +
  facet_wrap(~virus_family) +
  ylim(-10,max(t$data$y)+10)
ggsave("tree-figure/faceted-trees.pdf", width=11, height=8.5)

```